\documentclass[11pt]{article}
\usepackage{amsmath, amssymb}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{siunitx}

\title{Governing Equations for a Solar-Driven, Single-Bed PAM--LiCl SAWH Device}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Reference and scope}

This document describes the lumped model implemented in \texttt{solar\_lumped}, following Wilson \& D\'{i}az-Mar\'{i}n \textit{Device} (2025) Eqs.~1--6 and Note~S1 (COMSOL Table~S3 parameters). The device is a single hydrogel bed that alternates between open \emph{absorption} (night) and sealed \emph{desorption} (day) half-cycles. Thermal balances for the gel, solar absorber, and cover glass are solved as quasi-steady algebraic equations; the condenser temperature is transient during desorption.

\textbf{Subscripts:} $g$ = hydrogel; $abs$ = solar absorber; $gl$ = cover glass; $cond$ = aluminum condenser; $amb$ = ambient.

\section{State variables}

All quantities are per unit footprint area (\si{\metre\squared}) unless noted.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Symbol} & \textbf{Units} & \textbf{Role} \\
\midrule
$c_w$ & \si{\mol\per\cubic\metre} & Brine water concentration in gel (Eq.~5 state) \\
$H$ & \si{\metre} & Hydrogel thickness (Eq.~6 state; reference $H_0$) \\
$T_{cond}$ & \si{\celsius} & Condenser temperature (transient during desorption only) \\
$T_g, T_{abs}, T_{gl}$ & \si{\celsius} & Quasi-steady thermal states (algebraic each step) \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Absorption phase ODEs:} $\dot{c}_w$, $\dot{H}$ with $Q_{solar}=0$, $\dot{m}_{des}=0$, $\dot{T}_{cond}=0$.

\textbf{Desorption phase ODEs:} $\dot{c}_w$, $\dot{H}$, $\dot{T}_{cond}$ with $T_g$, $T_{abs}$, $T_{gl}$ from the coupled thermal solve.

\section{Thermal balances (Wilson Eqs.~1, 3, 4)}

During each integration step the gel, absorber, and glass temperatures satisfy the steady residual form (Eqs.~1, 3, 4 set to zero):

\begin{equation}
U_{gel}(T_{abs} - T_g)
= h_{conv,g}(T_g - T_{cond})
+ \dot{m}_{des}\,h_{des}
+ \varepsilon_{gc}\sigma(T_g^4 - T_{cond}^4)
\label{eq:gel}
\end{equation}

\begin{equation}
k_{air}\frac{T_{abs} - T_{gl}}{L_{ins}}
+ \varepsilon_{ag}\sigma(T_{abs}^4 - T_{gl}^4)
= h_{amb}(T_{gl} - T_{amb})
+ \sigma(T_{gl}^4 - T_{amb}^4)
\label{eq:glass}
\end{equation}

\begin{equation}
\varepsilon_{abs}\,\tau_{glass}\,Q_{solar}
= k_{air}\frac{T_{abs} - T_{gl}}{L_{ins}}
+ \varepsilon_{ag}\sigma(T_{abs}^4 - T_{gl}^4)
+ U_{gel}(T_{abs} - T_g)
\label{eq:absorber}
\end{equation}

\textbf{Correlations (Note~S1):}
\begin{itemize}
\item $h_{conv,g}$: Hollands et al.\ natural convection between parallel plates in the vapor gap $L_g - H$ (tilt-dependent).
\item $\varepsilon_{gc} = 1/(1/\varepsilon_{gel} + 1/\varepsilon_{Al} - 1)$: parallel-plate radiative exchange.
\item $U_{gel} = k_{Al}/L_{Al}$: gel--absorber conduction (Table~S3).
\item Absorption: $Q_{solar}=0$; desorption: $Q_{solar}\ge 0$ from weather profile.
\end{itemize}

\section{Condenser energy balance (Wilson Eq.~2, transient)}

During desorption only:
\begin{equation}
(\rho c_p L)_{cond}\,\frac{dT_{cond}}{dt}
= h_{conv,g}(T_g - T_{cond})
+ \dot{m}_{des}\,h_{fg}
+ \varepsilon_{gc}\sigma(T_g^4 - T_{cond}^4)
- A_r\,h_{amb}(T_{cond} - T_{amb})
\label{eq:condenser}
\end{equation}

where $A_r$ is the fin area ratio and $h_{conv,cond} = A_r\,h_{amb}$.

\section{Mass transport (Wilson Eqs.~5--6)}

\begin{equation}
\frac{dc_w}{dt}
= \frac{g}{H_0}\,\frac{P_{sat}(T_g)}{RT_g}\,(C_R - a_w)
\label{eq:dcw}
\end{equation}

\begin{equation}
\frac{dH}{dt}
= \frac{MW_{w}}{\rho_{sol}}\,\frac{P_{sat}(T_g)}{RT_g}\,(C_R - a_w)
\label{eq:dH}
\end{equation}

\textbf{Concentration ratio $C_R$:}
\begin{align}
C_R &= RH_{amb} && \text{(absorption, open chamber)} \\
C_R &= \frac{P_{sat}(T_{cond})}{P_{sat}(T_g)}\,\frac{T_g + 273.15}{T_{cond} + 273.15}
&& \text{(desorption, vapor gap)}
\end{align}

\textbf{Mass-transfer coefficient $g$:}
\begin{align}
g &= g_{chamber} && \text{(absorption)} \\
g &= h_{conv,g}\,\frac{D_{air}}{k_{air}} && \text{(desorption; Lewis-number analogy, Note~S1 Eq.~S5)}
\end{align}

\textbf{Water activity $a_w$ (PAM--LiCl):} brine activity from salt thermodynamics, capped by the measured DVS isotherm during absorption (Note~S2). During desorption, brine activity alone.

\textbf{Desorption mass flux from gel inventory:}
\begin{equation}
\dot{m}_{des} = MW_w\,\bigl(-\dot{c}_w\,H - c_w\,\dot{H}\bigr), \qquad \dot{m}_{des} \ge 0
\label{eq:mdes}
\end{equation}

\section{Thermal--mass coupling constraint}

During desorption, $\dot{m}_{des}$ in Eq.~\eqref{eq:gel} must equal the flux implied by Eqs.~\eqref{eq:dcw}--\eqref{eq:mdes}. The implementation finds $\dot{m}_{des}$ as the root of $f(\dot{m}_{des}) = \dot{m}_{calc}(\dot{m}_{des}) - \dot{m}_{des} = 0$ (Brent's method), ensuring Eqs.~1 and 5--6 are self-consistent at each ODE evaluation.

\section{Constraints and cycle structure}

\begin{itemize}
\item $c_w \in [c_{w,\min}, c_{w,\max}]$; rates clipped at bounds.
\item $H \ge H_0$; swelling allowed in absorption ($\dot{H}\ge 0$ at $H=H_0$); no shrinkage below $H_0$ in desorption.
\item Desorption: enforce $\dot{c}_w \le 0$, $\dot{H} \le 0$.
\item Daily cycle: absorption half-cycle ($\sim$12~h, no solar) followed by desorption half-cycle ($\sim$12~h, solar + condenser transient).
\item Initial $c_w$ from fabrication equilibrium RH (DVS isotherm at $H_0$) unless overridden.
\item Thermal efficiency: $\eta_{th} = \dot{m}_{water}\,h_{fg} / \int Q_{solar}\,dt$ over the desorption phase.
\end{itemize}

\section{Default parameters (Table~S3 / \texttt{table\_s3.py})}

\begin{table}[h]
\centering
\small
\sisetup{table-format=3.3}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Parameter} & \textbf{Symbol} & \textbf{Default} \\
\midrule
Hydrogel reference thickness & $H_0$ & \SI{4}{\milli\metre} \\
Vapor gap length & $L_g$ & \SI{40}{\milli\metre} \\
Insulation gap & $L_{ins}$ & \SI{5}{\milli\metre} \\
Aluminum plate thickness & $L_{Al}$ & \SI{0.125}{\inch} \\
Chamber convection (abs.) & $g_{chamber}$ & \SI{0.0085}{\metre\per\second} \\
Solution density & $\rho_{sol}$ & \SI{1250.2}{\kilogram\per\cubic\metre} \\
Desorption enthalpy & $h_{des}$ & \SI{2.32e6}{\joule\per\kilogram} \\
Condensation enthalpy & $h_{fg}$ & \SI{2.256e6}{\joule\per\kilogram} \\
Gel emissivity & $\varepsilon_{gel}$ & 1.0 \\
Condenser emissivity & $\varepsilon_{Al}$ & 0.05 \\
Absorber emissivity & $\varepsilon_{abs}$ & 0.95 \\
Glass transmittance & $\tau_{glass}$ & 0.9 \\
Tilt angle & --- & \SI{30}{\degree} (baseline) \\
Fin area ratio & $A_r$ & 7.1 (baseline) \\
Salt & --- & LiCl, salt:polymer = 4:1 \\
\bottomrule
\end{tabular}
\end{table}

\section*{Integration}

The coupled system is integrated with SciPy \texttt{solve\_ivp} (Radau, stiff). State during desorption: $\mathbf{y} = [c_w,\, H,\, T_{cond}]$; during absorption: $\mathbf{y} = [c_w,\, H]$.

\end{document}
